Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-07-18 paper/manuscript.tex
update subsampling.

Some of the authors' experience has shown that significantly more reliable estimators at lower numbers of samples are obtained by performing bootstrap sampling on data that has been uncorrelated. We do not provide a full tutorial of bootstrapping here, but the basic idea is that one takes the $N_{sample}$ data points collected and draws, with replacement, $N_{sample}$ samples.  This means some of the data points will be represented multiple times, and other data points none. This process is then repeated $N_{bootstrap}$ times, where $N$ should generally be at least 50 to up to 200 for good estimates of the standard error of the estimator.  The data is then \textit{treated} as $N_{bootstrap}$ samples from the true distribution, and statistics such as the standard error are computed from this artificial set of $N_{bootstrap}$ experiments.  In the limit of $N_{sample}$ large, it can be shown that the bootstrap estimate of the standard error converges to the true population standard error.  One way this can be seen is that bootstrap estimators of the data are independent samples of the \textit{actual} distribution of the experiment performed, i.e., a bunch of $\delta$ functions at $N_{sample}$ points, rather than being independent samples from the true, usually continuous distribution.  As more and more data is collected, our collection of $\delta$ functions becomes more and more like the true distribution. Another advantage of bootstrap sampling is that one can construct decent estimates of error bounds with arbitrary percentiles, and not just standard errors of the estimates, which assume Gaussian behavior of the underlying error, and that the error results are Gaussian as well.

One important point is that the bootstrap samples use to estimate the standard error should be decorrelated, or else the bootstrap estimate of the standard error will be artificially small.  More discussion of subsampling and decorrelation is included in section~\ref{sec:decorrelating-samples}. Another alternative to bootstrap sampling of individual points that better captures the correlations in the data is block bootstrapping~\cite{???}.  Block bootstrapping is somewhat more complicated and less common to perform and analyze than bootstrap estimates of the error, so we will not go into details, but it can also provide more reliable estimates of uncertainty than the estimators above.

@@

@@

We recommend using the robust statistical inefficiency computation routines available within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages.

%Cooment: Finlay, can you add description/citation of RED?

@@

However, samples uncorrelated by this procedure are only approximately uncorrelated.  Recent careful investigation has shown that the lowest uncertainty results can be obtained by using \textit{all} of the data to obtain an estimate of the free energy itself, but using only subsampled data to calculate error estimates using the formulas or by using bootstrapping.  In fact, one can show that the standard deviation of the estimate determined by data that has been subsampled is the unbiased estimator of the free energy estimate using all the data, but is actually an underestimate of the uncertainty of estimate
using just the subsampled data.

We don't need to use all of the data to recover the remaining correlation.  In practice, if one subsamples the data with a
correlation interval of $g/10$, where $g$ is the correlation time. Thus, if $g=3.0$ in units of timesteps, it is best to use all of the data.  If $g=100.0$, then one loses essentially no information by subsampling with $g=10.0$, and still saves significant time by not using all of the now-completely duplicative data.

@@

Our overall recommendations are therefore to determine the subsampling period $g$, subsample data at $g/10$ (i.e. more frequently than specified by $g$) to compute the free energy estimates, and then use fully subsampled data (with subsampling interval $g$) to estimate the uncertainties of the estimates.

>>>>

Some of the authors' experience has shown that significantly more reliable estimators at lower numbers of samples are obtained by performing bootstrap sampling on data that has been uncorrelated. We do not provide a full tutorial of bootstrapping here, but the basic idea is that one takes the $N_{sample}$ data points collected and draws, with replacement, $N_{sample}$ samples.  This means some of the data points will be represented multiple times, and other data points none. This process is then repeated $N_{bootstrap}$ times, where $N$ should generally be at least 50 to up to 200 for good estimates of the standard error of the estimator.  The data is then \textit{treated} as $N_{bootstrap}$ samples from the true distribution, and statistics such as the standard error are computed from this artificial set of $N_{bootstrap}$ experiments.  In the limit of $N_{sample}$ large, it can be shown that the bootstrap estimate of the standard error converges to the true population standard error.  One way this can be seen is that bootstrap estimators of the data are independent samples of the \textit{actual} distribution of the experiment performed, i.e., a bunch of $\delta$ functions at $N_{sample}$ points, rather than being independent samples from the true, usually continuous distribution.  As more and more data is collected, our collection of $\delta$ functions becomes more and more like the true distribution. Another advantage of bootstrap sampling is that one can construct decent estimates of error bounds with arbitrary percentiles, and not just standard errors of the estimates, which assume Gaussian behavior of the underlying error, and that the error results are Gaussian as well.  The bootstrap approach, on the other hand, captures non-Gaussian behavior in the data.

One important point is that the bootstrap samples use to estimate the standard error should be decorrelated, or else the bootstrap estimate of the standard error will be artificially small.  More discussion of subsampling and decorrelation is included in section~\ref{sec:decorrelating-samples}. Another alternative to bootstrap sampling of individual points that better captures the correlations in the data is block bootstrapping~\cite{???}.  Block bootstrapping is somewhat more complicated and less common to perform and analyze than bootstrap estimates of the error, so we refer readers to the citations for more description, but it can also provide more reliable estimates of uncertainty than the estimators above.

@@

%MRS: Finlay, you can add information about RED here.

@@

We recommend using the robust statistical inefficiency computation routines available within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages, though we caution that any such algorithms will underestimate the correlation time with insufficient data. Such approaches also assume that the data doesn't have any long time scale drift over the course of the simulation, which is a clear sign of lack of equilibration.  Generally, most methods will give reasonable correlation times when the simulation lasts 40--50 correlation times,

@@

However, samples uncorrelated by this procedure are only approximately uncorrelated.  Recent careful investigation has shown that the lowest uncertainty results can be obtained by using \textit{all} of the data to obtain an estimate of the free energy itself, but using only subsampled data to calculate error estimates using the formulas or by using bootstrapping.
In fact, one can show that the standard deviation of the estimate determined by data that has been subsampled is the unbiased estimator of the free energy estimate using all the data, but is actually an underestimate of the uncertainty of estimate using just the subsampled data.
%MRS: Finlay, is there a citation for the mathematical analysis you showed in the issue tracker?  This would be a good place to it.

We don't need to use all of the data to recover the remaining correlation.  In practice, if one subsamples the data with a
correlation interval of $g/10$, where $g$ is the correlation time, this will recover the missing correlation.  Thus, if $g=3.0$ in units of timesteps, it is best to use all of the data.  If $g=100.0$, then one loses essentially no information by subsampling with $g=10.0$ in estimating the free energies, and still saves significant time by not using all of the now-completely duplicative data.  Note that subsampling with correlation time $g$ should still be used for estimating the uncertainties, perhaps requiring two passes of the simulation analysis code.

@@

Our overall recommendations are therefore to determine the subsampling period $g$, either use all of the data or subsample with something like $g/10$ (i.e. more frequently than specified by $g$) to compute the free energy estimates, and then use fully subsampled data (with subsampling interval $g$) to estimate the uncertainties of the estimates.  Uncertainties calculated with the full data set will be significant underestimates if $g > 1$.
